---
output: 
  pdf_document:
    number_sections: true
    keep_tex: true
classoption: landscape
geometry: "left=3.8cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
papersize: letter
header-includes:
  \usepackage{amsmath}
  \usepackage{pdflscape}
  \usepackage{tikz}
  \usepackage{xcolor}

  \newcommand{\blandscape}{\begin{landscape}}
  \newcommand{\elandscape}{\end{landscape}}

  \newcommand{\bcenter}{\begin{center}}
  \newcommand{\ecenter}{\end{center}}



# bibliography: bibliography.bib
# csl: vancouver-imperial-college-london.csl
---

```{r include = FALSE}
library(ggplot2)
library(tidyverse)
library(dplyr)
library(plyr)
library(readr)
library(tikzDevice)

data_all <- list.files(path = ".",     # Identify all csv files in folder
                       pattern = "*.out", full.names = TRUE) %>% 
  lapply(read_csv) %>%                                            # Store all files in list
  bind_rows                                                       # Combine data sets into one data set
```





```{r include = TRUE, echo=FALSE}
print(data_all)

  uniqModels <- unique(data_all[,1])

```



\pagenumbering{roman}     <!-- first page with Roman numbering -->

\newpage                  <!-- new page -->

\bcenter         % <!-- center text -->

\LARGE{My Report or Thesis Title} % <!-- make large text -->

\bigskip                % <!-- blank lines -->
\bigskip

\large{Dr Hale}

\ecenter              <!-- end center -->

\newpage 

\bcenter 

\large{Abstract}

\ecenter

\bigskip

...this is the abstract text...

\newpage 

\tableofcontents

\newpage
# List of tables{-}
\renewcommand{\listtablename}{} <!-- removes default section name -->
\listoftables

\newpage
# List of figures{-}
\renewcommand{\listfigurename}{}
\listoffigures

\newpage

\pagenumbering{arabic}

# Introduction{#intro}

# Methods{#meths}
## Equations
This is an $x=y$ equation.

## Bullets

# Results{#res}
## R-code chunk
## R-code inline
## Images
## Figures{#resultsFigs}

<!-- fig.cap=paste(uniqModels,"\\label{fig:figLand}") -->
 <!-- eval.after='fig.cap', -->

<!-- Figure \ref{fig:figLand} is on a landscape page. -->


```{r, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.height=5.2, fig.width=9.5, fig.align="center", dev="tikz", sanitize=TRUE, fig.cap = "Smooth Exp (Double)"}
data_all = filter (data_all, calculator != "DCT+Bessel+IDCT")
data_double = filter(data_all, bytes == 8)
data_float = filter(data_all, bytes == 4)
ggplot(data=data_double[data_double$function_name == "Smooth_exp",],aes( N,max_error, color=calculator,linetype=calculator))+geom_line()+geom_point() + scale_y_continuous(trans="log10", breaks = c(.1,.01,.001,1e-4,1e-5,1e-6,1e-7,1e-8,1e-9,1e-10,1e-11)) + scale_x_continuous(trans="log2",breaks=c(2,4,8,16,32,64,128,256)) 
```


```{r, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.height=5.2, fig.width=9.5, fig.align="center", dev="tikz",  sanitize=TRUE, fig.cap = "Smooth Poly"}
ggplot(data=data_double[data_double$function_name == "Smooth_poly",],aes( N,max_error, color=calculator,linetype=calculator))+geom_line()+geom_point() + scale_y_continuous(trans="log10", breaks = c(.1,.01,.001,1e-4,1e-5,1e-6,1e-7,1e-8,1e-9,1e-10,1e-11)) + scale_x_continuous(trans="log2",breaks=c(2,4,8,16,32,64,128,256)) 
```


```{r, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.height=5.2, fig.width=9.5, fig.align="center", dev="tikz", sanitize=TRUE, fig.cap = "Smooth Runge"}
ggplot(data=data_double[data_double$function_name == "Smooth_runge",],aes( N,max_error, color=calculator,linetype=calculator))+geom_line()+geom_point() + scale_y_continuous(trans="log10", breaks = c(.1,.01,.001,1e-4,1e-5,1e-6,1e-7,1e-8,1e-9,1e-10,1e-11)) + scale_x_continuous(trans="log2",breaks=c(2,4,8,16,32,64,128,256))
```

```{r, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.height=5.2, fig.width=9.5, fig.align="center", dev="tikz", sanitize=TRUE,fig.cap = "Nonsmooth Sqrt"}
ggplot(data=data_double[data_double$function_name == "Nonsmooth_sqrt",],aes( N,max_error, color=calculator,linetype=calculator))+geom_line()+geom_point() + scale_y_continuous(trans="log10", breaks = c(.1,.01,.001,1e-4,1e-5,1e-6,1e-7,1e-8,1e-9,1e-10,1e-11)) + scale_x_continuous(trans="log2",breaks=c(2,4,8,16,32,64,128,256)) 
```

```{r, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.height=5.2, fig.width=9.5, fig.align="center", dev="tikz", sanitize=TRUE, fig.cap = "Nonsmooth Abs"}
ggplot(data=data_double[data_double$function_name == "Nonsmooth_abs",],aes( N,max_error, color=calculator,linetype=calculator))+geom_line()+geom_point() + scale_y_continuous(trans="log10", breaks = c(.1,.01,.001,1e-4,1e-5,1e-6,1e-7,1e-8,1e-9,1e-10,1e-11)) + scale_x_continuous(trans="log2",breaks=c(2,4,8,16,32,64,128,256)) 
```

```{r, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.height=5.2, fig.width=9.5, fig.align="center", dev="tikz", sanitize=TRUE, fig.cap = "NonSmooth Runge Abs"}
ggplot(data=data_double[data_double$function_name == "Nonsmooth_runge_abs",],aes( N,max_error, color=calculator,linetype=calculator))+geom_line()+geom_point() +scale_y_continuous(trans="log10", breaks = c(.1,.01,.001,1e-4,1e-5,1e-6,1e-7,1e-8,1e-9,1e-10,1e-11)) + scale_x_continuous(trans="log2",breaks=c(2,4,8,16,32,64,128,256)) 
```



#}

## Tables{#resultsTables}

# Conclusion{#con}
We described methods in Section \ref{meths} and plotted the results in Section \ref{resultsFigs}. 
Supplementary material is in Appendix \ref{app:mat}.

# References{-}
<div id="refs"></div>

\renewcommand{\thesection}{\Alph{section}}
\setcounter{section}{1}

# Appendix{-}
## Supporting material{#app:mat}
## Supporting code